Modalities for Location Virtualization

### 2022-09-17 background.tex
separted stuff

\section{Background}
\label{sec:background}
\subsection{Virtual Memory Managers}
\label{sec:backgroundonvmm}
\subsection{Modal Logic}
\label{sec:backgroundonmodallogic}
Work on modal logic extends back for decades with many widely-varied variations; we discuss here only the pieces most relevant to our work.
Broadly speaking modal logics incorporate \emph{modal operators}, which take as arguments a proposition expected to be true in another time~\cite{pnueli1977temporal}, place~\cite{gordon2019modal,goranko1996hierarchies,areces2001hybrid,gargov1993modal}, or circumstance~\cite{hintikka1962knowledge,halpern1985guide}, and result in a proposition true in the \emph{current} time, place, or circumstance at which the truth of the use of the modal operator is being evaluated. Classic examples include modal necessity $\square P$ describing that $P$ is \emph{necessarily} true, $\mathsf{G}~P$ meaning $P$ is true \emph{globally} (i.e., forever from this time onwards), or $K_i(P)$ describing that a particular participant $i$ \emph{knows} that $P$ is true. The latter is an example of \emph{multimodal} logic, where there is an indexed family of modalities (modal operators) parameterized by some dimension of interest (there, participants).
A closely related variant of multimodal logic is \emph{dynamic logic}~\cite{pratt1976semantical}, a logic of weakest preconditions~\cite{dijkstra-75} which works with modalities of the form $[p](P)$, which states that \emph{in the current program state}, \emph{if} program $p$ is run then afterwards $P$ will hold (modulo non-termination).
This same idea is used to encode Hoare triples in the Iris program logic~\cite{krebbers2017essence}, using the same encoding as in Pratt's original presentation, where a Hoare triple $\{P\}C\{Q\}$ is encoded as $P\rightarrow[C](Q)$. Unlike classic work on dynamic logic~\cite{harel2000dynamic}, Iris applies these ideas in a \emph{substructural} setting (separation logic) where distributivity laws over substructural connectives must be considered. For both the dynamic modality and the later modality $\blacktriangleright$ used for guarded recursive predicates, these modalities satisfy axioms of the form $M(P)\ast Q\vdash M(P\ast Q)$, but not the reverse.
Iris is not the first combination of modal and substructural logic~\cite{dovsen1992modal,restall1993modalities,d1997grafting,kamide2002kripke,licata2017fibrational}, but is certainly the most heavily tested.
A hallmark of a unary logical operator $M$ being a modality is if $M$ satisfies a property akin to $(\upvarphi\rightarrow\psi)\rightarrow M(\upvarphi)\rightarrow M(\psi)$, which roughly states that modus ponens holds under the modality.\footnote{Afficionados of modal logic will note that this property is not quite Axiom K (which requires the initial implication to also hold under $M$), but follows from K and a necessitation rule $\upvarphi\rightarrow M(\upvarphi)$. Non-necessitive modalities typically satisfy the weaker property we call out above.}
% Most pertinent to our work is that Iris itself has a modalities built into the logic to solve problems with step-indexing / guardedness (the later modality) and general infrastructure for duplicable assumptions / resources (the persistent modality, reminiscent of linear logic's exponential). And of course the central concept in the use of Iris as a program logic is the use of the weakest precondition modality, which like our modalities is defined \emph{within} the logic itself, rather than being part of the base logic.

Another pillar of our work is hybrid logic, a branch of modal logic with \emph{names} (called \emph{nominals}) for states in Kripke models~\cite{blackburn1995hybrid,goranko1996hierarchies,areces2001hybrid,gargov1993modal}, in contrast to the typical flavor of modality which refers to an \emph{unspecified} (in the assertion) set of alternative circumstances.

Our work draws on ideas from hybrid modal logics, applied in the context of Iris's higher-order separation logic~\cite{jung2018iris}. We define our modalities and prove their rules directly within Iris, so there is no model-theoretic novelty in our work. Instead, we show that targeted use of modalities combining the ideas of \emph{named resources} with the power of substructural reasoning enable clear specification of general reasoning principles for data structures. Along the way we take advantage of the fact that modalities need not be injections between the \emph{same} logics, but in fact the propositions that modal operators contextualize can come from \emph{other} logics; in our case we exploit the fact that Iris itself is parameterized by a choice of step-indexed~\cite{ahmed-appel-virga-02} \textsf{BI}-algebra~\cite{ohearn1999bunched} to allow convenient specifications in embedded data-structure-relative logics.

>>>>

\section{Background}
\label{sec:background}
\subsection{Machine Model}
\label{sec:backgroundonmachinemodel}
\mytodo{Colin's part}
\subsection{Virtual Memory Managers}
\mytodo{Colin's part}
\label{sec:backgroundonvmm}
\subsection{Modal Logic}
\label{sec:backgroundonmodallogic}
\mytodo{Colin's part}

### 2022-09-17 intro.tex
separted stuff

\section{Introduction}
\label{sec:intro}
In the UNIX-inspired designs typical of modern general-purpose kernels [93, 82, 89, 91], virtual memory management, the filesystem, and device subsystem are tightly
coupled, rely upon other core OS services (scheduling, interrupt dispatch, etc.), and are central
to higher-level functionality. For example, in Solaris Internals [91], McDougall and Mauro write:
“The virtual memory sub-system can be considered the core of a Solaris instance, and the implementation of Solaris virtual memory affects just about every other subsystem in the operating
system.” So by verifying a kernel with full-featured versions of these essential systems, following the
designs kernel developers have found valuable enough to preserve for decades, we can make essential
progress towards handling real kernels written using these designs. This goal is feasible because
the verification challenges faced by these systems are all instances of just three core concepts from
systems research and development, which apply more broadly than just in OS kernels: location
virtualization, caching, and using asymmetric protocols to coordinate shared resource usage. We
believe all three concepts are amenable to direct treatment using tools from modal logic.

Virtual memory managers (VMMs) use hardware memory management units (MMUs) on each
CPU to virtualize the addresses of memory regions. Software actually accesses memory in terms of
a logical (virtual) address which is translated to a physical address in memory by the MMU based
on kernel-controlled page tables stored in memory: a hardware-defined sparse tree-map structure,
which itself is accessed by the kernel through virtual addresses. This tree structure may alias the
same physical memory to multiple virtual addresses. Whole trees are shared between threads of a
process, and subtrees or data pages may be shared across processes (e.g., the kernel mapping ~\cite{91,
93}). Accessing an unmapped address produces a fault which the OS can use to kill the process
or take corrective action (like add a mapping) and resume the interrupted code. This allows the
kernel to restrict which memory is accessible to different processes, hide memory limits from user
programs, ensure process isolation, implement paging and copy-on-write behaviors for performance
and resource controls, and even implement live migration of virtual machines in hypervisors ~\cite{110,
31}. Modern hardware extends this idea with an IOMMU ~\cite{71, 11} the OS can use to similarly
mediate hardware device access to memory.

We plan to use families of modalities to describe
the truth of assertions in the virtual resources corresponding to virtual address spaces or logical
locations in a file system: a modality for each concept, with corresponding assertions about the
underlying physical resource.
For virtual memory, this might correspond to an assertion [r]P indicating that P holds in the
virtual address space rooted at r. This is essentially what hybrid logic [19, 49, 10, 48] calls a
satisfaction operator, which evaluates the truth of a predicate in a named alternative state (here,
address space). An example usage of this would be as part of the invariant for a sleeping thread
— ∃R : registerset. Context(ctxt, R) ∗ ∃P : Prop. [R.cr3]P ∗ {P ∗ CurrentRegs(R)}(R.rip) — that the
context structure at virtual address ctxt holds a certain register set, and that the precondition for
the stored instruction pointer rip is that register set and the virtual invariant P, which is true in
the saved context’s address space rooted at cr3. This would be required to ensure safety of context
switches between address spaces.
Virtual points-to assertions would need to be defined in terms of physical points-to assertions
and hardware-specific address translation assertions (as in Kolanski and Klein’s work), and paged
points-to assertions indicating the VMM had committed swap space to hold the appropriate value
on disk. The semantics of address translation assertions would be given as an extension to Kolanski
and Klein’s approach, which implicitly treats the memory state as a modal frame and the choice of
page table root as a world in Kripke-style semantics. They stopped short, however, of incorporating
the choice of world into the logic’s assertions, or treating overlapping address spaces, or transitions

between address spaces. Overlapping address spaces — both shared data, and shared page tables
which may see updates — can be coordinated with protocols to ensure those updates preserve
address space invariants. For filesystem addressing modalities, the interpretation is dictated by the
entirely by filesystem design rather than the hardware, but the concept is similar.
Both systems have an implicit notion of a “current” world. CPUs have a current address space
based on their page table register (CR3 for x86-64), and the logic must support reasoning about
changing the current address space. Copy-on-write filesystems have a corresponding fixed location
on disk pointing to the physical block of the latest filesystem snapshot. For example, for CPUs, we
would ideally like proof rules close to the following:
{P ∗ CR3 = r1 ∗ [r2](Q) ∗ rax = r2}mov cr3 rax{[r1](P) ∗ CR3 = r2 ∗ Q ∗ rax = r2}
This rule captures the notion that switching address spaces pulls information out of one modality
into the “current view” of memory, and leaves other assertions true relative to the previous address
space. But real code requires stronger reasoning. Code that performs task switches typically
establishes an invariant in one address space — such as that the save-context of the thread being
taken off the CPU is valid — that must still be known true in the new address space. The right
way to expose this knowledge transfer is not obvious

>>>>

\section{Introduction}
\label{sec:intro}
\mytodo{Colin's Part}
